---
title: "**Classifying Chronic Kidney Disease Using a Multivariate Binary Logistic Regression Model**"
author: "Aanish Pradhan"
date: \today
output: 
  pdf_document: 
    keep_tex: yes
    citation_package: biblatex
biblatexoptions: [backend=biber, citestyle=numeric, bibstyle=numeric, autocite=superscript]
bibliography: references.bib
---

\setcounter{secnumdepth}{5}
\setcounter{tocdepth}{5}

# Abstract

Chronic Kidney Disease (CKD) is one of the leading causes of death in the 
United States. In 2020, renal diseases accounted for approximately 53,000 
deaths. It is estimated that approximately 15% of adults in the U.S. have CKD.
Furthermore, of the 15% of adults that have CKD, 90% of them are not aware that 
they have the condition [@CDC2021]. CKD is a challenging condition for 
physicians to diagnose and often requires a combination of laboratory testing 
and physical examinations to be able to diagnose a patient. We attempted to 
construct a statistical classification model that could determine whether or 
not a patient has CKD from a set of various predictors that could be obtained 
from laboratory blood test results or a physical examination. Our approach 
consisted of constructing and training several multivariate binary logistic 
regression models using various feature selection algorithms. The resulting 
models were benchmarked on a batch of testing data set aside earlier. The 
optimal model was chosen based on which model demonstrated the most favorable 
results in a confusion matrix. All models were able to obtain an accuracy of 
over 93%.

# Introduction

Chronic Kidney Disease is an umbrella phrase, used to refer to a multitude of 
chronic, degenerative (i.e. loss of kidney function over time) renal disorders 
[@Versino2019]. Diagnosing CKD is problematic because its symptoms do not 
present until later in life when the condition has seriously progressed. For 
this reason, CKD is often called a "silent killer" [@Kopyt2006]. Furthermore, 
CKD presents with similar symptoms as acute kidney injury as well as 
completely unrelated conditions. For example, microscopic hematuria (presence 
of red blood cells in the urine) and proteinuria (presence of protein in the 
urine) are characteristic symptoms of kidney disease. However, they are also 
observed in individuals after strenuous exercise such as long-distance running 
and weightlifting. Physicians have to use a combination of metrics collected 
over time from full-body physical examinations and laboratory blood test 
results as well as their intuition to diagnose CKD in patients.

# Methods

Our approach consists of three phases: an initial setup, modeling and, lastly, 
testing.

## Setup

In our setup phase, we will acquire, clean, format and conduct some initial 
exploratory data analysis (EDA).

### Data Collection

Our data will come from a dataset housed in the University of 
California-Irvine's Machine Learning Repository available [\underline{here}](https://archive.ics.uci.edu/ml/machine-learning-databases/00336/Chronic_Kidney_Disease.rar) [@Dua2019]. The data itself was collected in a study conducted over the 
span of two months at the Alagappa University Health Care Centre in Tamilnadu, 
India. No other details were given regarding collection methods.

```{r Data Collection}
# Read original dataset
originalData <- read.csv("Data/chronic_kidney_disease_full.csv", header = TRUE)
colnames(originalData)
```

The dataset contains a multitude of features such as age, blood pressure, serum 
creatinine and other biometrics that are obtained from full-body physical exams 
and laboratory blood tests.

### Data Wrangling

```{r Data Wrangling Extraneous Characters}
head(originalData, n = 5)
```

We observe some extraneous characters contaminating the dataset. We will 
replace extraneous characters, whitespace and blankspace with "NA" values.

```{r Data Wrangling Replacement}
# Replace extraneous characters, whitespace and blankspace with NA values
replacedData <- read.csv("Data/chronic_kidney_disease_full.csv", 
	header = TRUE, na.strings = c("", " ", "?"))
head(replacedData, n = 1)
```

### Data Cleaning

```{r Data Cleaning Missing}
any(is.na(replacedData))
```

Our dataset contains observations with missing values. We will discard these 
observations.

```{r}
# Omit observations with NA entries
cleanedData <- na.omit(replacedData)
any(is.na(cleanedData))
```

### Data Formatting

```{r Data Formatting Type Check}
str(cleanedData)
```

Some of our features were read in with the wrong type. We will correct the type 
of the features.

```{r Data Formatting Type Correction}
# Correct the variable types
formattedData <- cleanedData
formattedData$id <- as.integer(formattedData$id)
formattedData$X.sg. <- as.factor(formattedData$X.sg.)
formattedData$X.al. <- as.factor(formattedData$X.al.)
formattedData$X.su. <- as.factor(formattedData$X.su.)
formattedData$X.rbc. <- as.factor(formattedData$X.rbc.)
formattedData$X.pc. <- as.factor(formattedData$X.pc.)
formattedData$X.pcc. <- as.factor(formattedData$X.pcc.)
formattedData$X.ba. <- as.factor(formattedData$X.ba.)
formattedData$X.htn. <- as.factor(formattedData$X.htn.)
formattedData$X.dm. <- as.factor(formattedData$X.dm.)
formattedData$X.cad. <- as.factor(formattedData$X.cad.)
formattedData$X.appet. <- as.factor(formattedData$X.appet.)
formattedData$X.pe. <- as.factor(formattedData$X.pe.)
formattedData$X.ane. <- as.factor(formattedData$X.ane.)
formattedData$X.class. <- as.factor(formattedData$X.class.)
str(formattedData)
```

With our dataset cleaned, we can proceed with exploratory data analysis.

### Exploratory Data Analysis

We can begin EDA with a 5-number summary of the features in our prepared data.

```{r EDA Dataframe Summary}
summary(formattedData)
```

We have several continuous numerical variables. We will examine correlation 
between the features.

```{r EDA Correlation Matrix Scatterplot}
# Extract continuous numerical variables
correlationMatrix <- data.frame(formattedData$X.age., formattedData$X.bp., 
	formattedData$X.bgr., formattedData$X.bu., formattedData$X.sc., 
		formattedData$X.sod., formattedData$X.pot., formattedData$X.hemo., 
			formattedData$X.pcv., formattedData$X.wbcc., formattedData$X.rbcc.)

# Abbreviate names
colnames(correlationMatrix) <- c("X.age.", "X.bp.", "X.bgr.", "X.bu.", "X.sc.", 
	"X.sod.", "X.pot.", "X.hemo.", "X.pcv.", "X.wbcc.", "X.rbcc")

# Correlation plot
plot(correlationMatrix, main = "Scatterplot of Correlation Matrix")
```

The correlation matrix scatterplot shows some features are correlated with 
each other. We can quantify the correlation by examining the Pearson 
correlation coefficients computed from the correlation matrix.

```{r EDA Correlation Matrix}
# Compute Pearson correlation coefficients and round r-values
round(cor(correlationMatrix, method = "pearson"), digits = 2)
```

Some variables appear to have strong, linear relationships with other 
variables. This indicates that we could observe issues with multicollinearity 
in our models. Using a cutoff of $r = \pm 0.7$, we can identify which variables 
are highly correlated with others.

```{r EDA Correlation Matrix Collinear Variables}
abs(cor(correlationMatrix, method = "pearson")) > 0.7
```

We observe that \textcolor{red}{\texttt{X.bu}}, 
\textcolor{red}{\texttt{X.hemo.}}, and \textcolor{red}{\texttt{X.rbcc.}} 
exhibit multicollinearity with several other variables.

## Modeling

In our modeling phase, we will perform a training and test dataset split, 
construct our models using various feature selection algorithms and perform 
various model diagnostics.

### Train-Test Data Split

Our variable of interest is \textcolor{red}{\texttt{X.class.}}.

```{r Modeling Response Variable}
table(formattedData$X.class.)
```

We will randomize the rows of our dataset and perform a 50-50 train-test data 
split. The training data will contain 22 "ckd"-classified observations 57 
"notckd"-classified observations. The testing data will contain 21 
"ckd"-classified observations and 57 "notckd"-classified observations. In order 
to maintain reproducibility, we will use a sample seed of "42".

```{r Modeling Row Randomization}
# Generate a row-randomized dataset
set.seed(42)
randomizedData <- formattedData[sample(nrow(formattedData)), ]
```

From our randomized dataset, we can copy the first 22 "ckd"-classified 
observations into the training dataset and the subsequent 21 
"notckd"-classified observations into the testing dataset.

```{r Modeling Data Split}
# Construct testing dataset
testingData <- randomizedData # Dataset is constructed by Complement Rule

# Construct training dataset
trainingData <- testingData[-c(1:157), ] # Copy column names & preserve type

for (i in 1:length(testingData$id)) # Extract first 22 CKD observations
{
	if ((testingData[i, ]$X.class. == "ckd") & 
		(sum(trainingData$X.class. == "ckd") < 22))
	{
		trainingData[nrow(trainingData) + 1, ] <- testingData[i, ]
		testingData <- testingData[-c(i), ]
	}
}

for (i in 1:length(testingData$id)) # Extract first 57 non-CKD observations 
{
	if ((testingData[i, ]$X.class. == "notckd") & 
		(sum(trainingData$X.class. == "notckd") < 57))
	{
		trainingData[nrow(trainingData) + 1, ] <- testingData[i, ]
		testingData <- testingData[-c(i), ]
	}
}

rm(i) # Clears the counter variable from the environment

# Reorder datasets
trainingData <- trainingData[order(trainingData$id), ]
testingData <- testingData[order(testingData$id), ]
```

With our training and testing datasets in place, we can proceed with building a 
model.

### Feature Selection

We will construct our models using the Forward Selection, Backward Elimination, 
Sequential Selection (Bidirectional Elimination), Least Absolute Shrinkage and 
Selection Operator (LASSO) and Ridge Regression algorithms.

#### Forward Selection Algorithm

To run the Forward Selection algorithm, we will construct a Null 
(intercept-only) model and a Full (all regressors) model. The algorithm will 
iteratively add regressors to the Null Model until it is no longer optimal to 
do so.

```{r Modeling Forward Selection, warning = FALSE}
# Generate the Null Model
nullModel <- glm(X.class. ~ 1, data = trainingData, family = "binomial")

# Generate the Full Model
fullModel <- glm(X.class. ~ X.age. + X.bp. + X.bgr. + X.bu. + X.sod. + X.pot. + 
	X.hemo. + X.pcv. + X.wbcc. + X.rbcc., data = trainingData, 
		family = "binomial")

# Run Forward Selection algorithm
forwardModel <- step(nullModel, direction = "forward", 
	scope = list(upper = fullModel, lower = ~1), trace = 0)
summary(forwardModel)
```

#### Backward Elimination Algorithm

The Backward Elimination algorithm will iteratively remove the least 
statistically significant regressor from the Full Model until it is no longer
optimal to do so.

```{r Modeling Backward Elimination, warning = FALSE}
# Run the Backward Elimination algorithm
backwardModel <- step(fullModel, direction = "backward", trace = 0)
summary(backwardModel)
```

#### Sequential Selection Algorithm

The Sequential Selection algorithm will iteratively either add or remove a
regressor at each iteration until it is no longer optimal to do so.

```{r Modeling Sequential Selection, warning = FALSE}
# Run Sequential Selection algorithm
sequentialModel <- step(nullModel, direction = "both", 
	scope = formula(fullModel), trace = 0)
summary(sequentialModel)
```

The Sequential Selection algorithm yields the same linear model as the Forward 
Model, thus we can ignore this algorithm's output.

#### LASSO Regression Algorithm

To run the LASSO Regression algorithm, we will utilize the 
\textcolor{blue}{\texttt{glmnet}} package [@Friedman2010]. We will perform a 
k-fold cross-validation to find a value of $\lambda$ that minimizes the Mean 
Squared Error (MSE). The algorithm will optimize a loss function that takes 
into account the sum of the absolute value of the regressors' coefficients. By 
doing so, it imposes a penalty on the optimization, causing the regressor 
coefficients to "shrink" towards zero, thereby minimizing the number of 
regressors required in the model.

```{r Modeling LASSO Cross-Validation, warning = FALSE}
# Load the glmnet package
library(glmnet)

# k-fold cross-validation
lassoY <- trainingData$X.class.
lassoX <- data.matrix(trainingData[, colnames(trainingData)[2:25]])
lassoCVModel <- cv.glmnet(lassoX, lassoY, alpha = 1, family = "binomial")
lassoBestLambda <- lassoCVModel$lambda.min
lassoBestLambda
```

```{r Modeling LASSO Regression}
# Run LASSO Regression algorithm
lassoModel <- glmnet(lassoX, lassoY, alpha = 1, lambda = lassoBestLambda, 
	family = "binomial")
coef(lassoModel)
```

#### Ridge Regression Algorithm

We will perform a k-fold cross-validation to find a value of $\lambda$ that 
minimizes the MSE. Similar to the LASSO Regression algorithm, the Ridge 
Regression algorithm will minimize a loss function that accounts for the 
coefficients of the regressors. However, the loss function for the Ridge 
Regression algorithm involves the sum of the squares of the coefficients 
regressors as opposed to the sum of the absolute values.

```{r Modeling Ridge Cross-Validation}
# k-fold cross-validation
ridgeY <- trainingData$X.class.
ridgeX <- data.matrix(trainingData[, colnames(trainingData)[2:25]])
ridgeCVModel <- cv.glmnet(ridgeX, ridgeY, alpha = 0, family = "binomial")
ridgeBestLambda <- ridgeCVModel$lambda.min
ridgeBestLambda
```

```{r Modeling Ridge Regression}
# Run Ridge Regression algorithm
ridgeModel <- glmnet(ridgeX, ridgeY, alpha = 0, lambda = ridgeBestLambda, 
	family = "binomial")
coef(ridgeModel)
```

### Model Checking

Before validating our models, we must check our assumptions.

1. **Binary Response**. Our dependent variable must be a categorical nominal 
variable with two levels.

```{r Assumptions Response}
table(formattedData$X.class.)
```

Our assumption is met.

2. **Independence of Observations**. Our observations need to be independent 
from one another. Intuitively, one patient being diagnosed with CKD does not 
conceivably influence whether or not another patient is diagnosed with CKD. The 
inverse of this statement also is reasonably (i.e., a patient being diagnosed 
as healthy (without CKD) does not influence another patient being diagnosed 
as healthy). Our assumption is met.

3. **Multicollinearity**. The regressors of our models should not exhibit high 
amounts of multicollinearity between each other. Because the Forward and 
Backwards Models were generated from non-penalizing regression methods, we will 
explicitly check for multicollinearity using Variance Inflation Factors (VIF) 
which can be computed from the \texttt{vif()} function in the 
\textcolor{blue}{\texttt{car}} package [@Fox2019].

```{r Assumptions Multicollinearity}
# Load car package
library(car)

# Multicollinearity Detection
vif(forwardModel)
vif(backwardModel)
```

The Forward Model exhibits VIF scores over 10 for both regressors, indicating a 
serious multicollinearity issue with the model. The Backward Model exhibits VIF 
scores under 5 for each regressor, indicating a acceptable amount of 
correlation between the regressors. We will abandon the Forward Model and 
retain the other models.

4. **Outliers**. Our model should not contain any extreme outliers or high 
influence points (HIP). We will check for outliers and HIPs using Cook's 
Distance and discard those observations from both models if they are found.

```{r Assumptions Outliers}
# Cook's Distance Plot Backward Model
plot(cooks.distance(backwardModel), main = "Cook's Distance Values in the 
	 Backward Model (Cutoff = 1", xlab = "Observation #", ylab = "Di Value")
```

There are no observations with a $D_{i}$ value greater than 1. Therefore, there 
are no HIPs in this model. Our assumption is met.

5. **Linearity Between Logit of Response and Regressor**. For each regressor in 
our model, there needs to be a linear relationship between the logit of the 
response and the explanatory variable. We will check for linearity by examining 
a scatterplot of the log-odds versus the regressor.


```{r Assumptions Linearity, warning=FALSE}
# Linearity Check for the Backward Model
backwardModelLogOdds <- backwardModel$linear.predictors
cor(trainingData$X.bgr., backwardModelLogOdds)
cor(trainingData$X.bu., backwardModelLogOdds)
cor(trainingData$X.wbcc., backwardModelLogOdds)


plot(trainingData$X.bgr., backwardModelLogOdds, 
	 main = "Backward Model Log Odds vs. Blood Glucose (X.bgr.) Scatterplot", 
		xlab = "Blood Glucose (mg/dL)", ylab = "Log Odds")
plot(trainingData$X.bu., backwardModelLogOdds,  
	 main = "Backward Model Log Odds vs. Blood Urea (X.bu.) Scatterplot", 
		xlab = "Blood Urea (mg/dL)", ylab = "Log Odds")
plot(trainingData$X.wbcc., backwardModelLogOdds,
	 main = "Backward Model Log Odds vs. White Blood Cell Count (X.wbcc.) 
		Scatterplot", xlab = "White Blood Cell Count (cells/mm^3)", 
			ylab = "Log Odds")
```

The Backward model's log odds have a strong negative linear relationship with 
the blood glucose and blood urea regressors. The model's log odds have a 
moderate negative linear relationship with the white blood cell count 
regressor. Our assumption is met for this model.

6. **Sample Size**. We require $\frac{(10 \times k)}{P(x)}$ number of 
observations where $k$ is the number of regressors and $P(x)$ is the 
expected probability of the least frequent outcome in the dataset.

```{r Assumptions Sample Size}
# Sample Size Check

table(trainingData$X.class.)

(10 * 4) / (22 / 79) # 4 regressors, 22 / 79 CKD-classified observations
```

Our training dataset contains only 79 observations. Thus our assumption will 
not be met. We will proceed anyways as our models have passed all other 
critera for logistic regression.

## Validation

Using our testing dataset, we will collect statistics such as sensitivity, 
specificity, positive predictive value and negative predictive value on the 
classification models.

### Backward Model

We will utilize the \texttt{%>%} (pipe) operator from the 
\textcolor{blue}{\texttt{dplyr}} package.

```{r Validation Backward Model, warning = FALSE}
# Load the dplyr package
library(dplyr)

# Run the Backward Model on the test dataset
backwardModelProbabilities <- backwardModel %>% predict(testingData, 
	type = "response")
backwardModelPredictedClasses <- ifelse(backwardModelProbabilities < 0.5, 
	"ckd", "notckd")
```

### LASSO Model

```{r Validation LASSO Model}
# Testing data matrix
testDataMatrix <- data.matrix(testingData[, 2:25])

# Run the LASSO Model on the test dataset
lassoModelPredictedProbabilities <- predict(lassoModel, s = lassoBestLambda, 
	newx = testDataMatrix, type = "response")
lassoModelPredictedClasses <- ifelse(lassoModelPredictedProbabilities < 0.5, 
	"ckd", "notckd")
```


### Ridge Regression Model

```{r Validation Ridge Regression Model}
ridgeModelPredictedProbabilities <- predict(ridgeModel, s = ridgeBestLambda, 
	newx = testDataMatrix, type = "response")
ridgeModelPredictedClasses <- ifelse(ridgeModelPredictedProbabilities < 0.5, 
	"ckd", "notckd")
```


# Results

We will use the \textcolor{blue}{\texttt{caret}} package to generate a 
confusion matrix for each of the models to collect sensitivity, specificity, 
positive predictive value (PPV) and negative predictive value (NPV) 
[@Kuhn2008]. 

```{r Results caret Package, warning = FALSE}
# Load the caret package
library(caret)
```

## Backward Model

```{r Backward Model Confusion Matrix}
confusionMatrix(table(backwardModelPredictedClasses, testingData$X.class.))
```

The Backward Model has an accuracy of 93.59%. It is able to correctly return a 
"ckd" classification, given that a patient does actually have CKD, 76.19% of the 
time and correctly return a "notckd" classification, given that a patient 
does not actually have CKD, 100% of the time. A patient has a 100% chance of 
having CKD given that the model returns a "ckd" classification for their 
biometrics and a 91.94% chance of not having CKD given that the model returns a 
"notckd" classification for their biometrics.

## LASSO Model

```{r LASSO Model Confusion Matrix}
confusionMatrix(table(lassoModelPredictedClasses, testingData$X.class.))
```

The LASSO Model has an accuracy of 100%. It is able to correctly return a 
"ckd" classification, given that a patient does actually have CKD, 100% of the 
time and correctly return a "notckd" classification, given that a patient 
does not actually have CKD, 100% of the time. A patient has a 100% chance of 
having CKD given that the model returns a "ckd" classification for their 
biometrics and a 100% chance of not having CKD given that the model returns a 
"notckd" classification for their biometrics.

## Ridge Regression Model

```{r Ridge Regression Model Confusion Matrix}
confusionMatrix(table(ridgeModelPredictedClasses, testingData$X.class.))
```

The Ridge Regresion Model has an accuracy of 100%. It is able to correctly 
return a "ckd" classification, given that a patient does actually have CKD, 
100% of the time and correctly return a "notckd" classification, given that a 
patient does not actually have CKD, 100% of the time. A patient has a 100% 
chance of having CKD given that the model returns a "ckd" classification for 
their biometrics and a 100% chance of not having CKD given that the model 
returns a "notckd" classification for their biometrics.

# Discussion & Conclusion

The most optimal model is the model that maximizes its accuracy, sensitivity, 
specificity, PPV and NPV. Based on this criteria, the LASSO and Ridge Regresion 
Models are the best models for classifying whether or not a patient has CKD. 
Additionally, the Backward Model is a strong model, demonstrating high 
accuracy, specificity, PPV and NPV. In the future, this analysis should be 
repeated using larger sample sizes. One potential re-approach to this study 
could involve the use of Generative Adversarial Neural Networks (GAN). GANs 
allow for the creation of new, but similar and useful data from random noise 
which can be used to handle issues involving low sample sizes [@Saxena2021].

\newpage
